---
title: "Assignment Three Project"
author: "Elphas Oyugi, Paul Hoos, Firuddin Aliyev"
date: "2025-06-14"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r Libraries}
library(sf)
library(ggplot2)
library(ggspatial)
library(dplyr)
library(rnaturalearth)
library(rnaturalearthdata)
library(gridExtra)
library(tidyr)
library(raster)
library(terra)
library(patchwork)
library(spgwr)
library(spData)
library(here)
```

#Loading Shapefile for Sachsen as a geographic region as an area of interest,
#Listing the files in the directory path folder

```{r}
shapefile_path <- here("data", "gadm41_DEU_shp", "gadm41_DEU_1.shp")
list.files(shapefile_path)
```

#Wrangling the shapefile data to execute analysis process, filtering data for Sachsen and visualization 

```{r}
print(shapefile_path)
germany_shp <- st_read(shapefile_path)
names(germany_shp)
unique(germany_shp$NAME_1)
Sachsen <- germany_shp %>% filter(NAME_1 == "Sachsen")
```
# Plotting Sachsen as a geographical area of interest and visualization

```{r}
plot(st_geometry(Sachsen), main = "Sachsen")
plot(st_geometry(Sachsen), col = "green", bg = "pink", main = "Sachsen")
```

#Descriptive summary of the shapefile data for Sachsen

```{r}
summary(Sachsen)
```
# Now its time to load Raster data for precipation and elevation, 
However,shapefile data also loaded again to filter Sachsen

```{r}
shapefile_path <- here("data", "gadm41_DEU_shp", "gadm41_DEU_1.shp")
germany_shp <- st_read(shapefile_path)
Sachsen <- germany_shp %>% filter(NAME_1 == "Sachsen")

raster_path <- here("data", "wc2.1_30s_elev.tif")
elev_raster <- rast(raster_path)


# Create file paths for all 12 months
months <- sprintf("%02d", 1:12)
precip_files <- file.path(here("data", "wc2.1_30s_prec"), paste0("wc2.1_30s_prec_", months, ".tif"))

# Load all into a raster stack
precip_stack <- rast(precip_files)  # creates a SpatRaster with 12 layers
```

#Build the sum of the months for annual precipation data (Crop and mask first for faster calculation)

```{r}
Sachsen <- st_transform(Sachsen, crs(precip_stack))
precip_stack_sachsen <- mask(crop(precip_stack, Sachsen), Sachsen)
precip_annual <- sum(precip_stack_sachsen)
plot(precip_annual, main = "Annual Precipitation in Sachsen")
```

# Now crop and mask raster elevation data to Sachsen only as a geographic region

```{r}
Sachsen <- st_transform(Sachsen, crs(elev_raster))
elev_sachsen <- mask(crop(elev_raster, Sachsen), Sachsen)
```

# Combine elevation and precipation

```{r}
elev_df <- as.data.frame(elev_sachsen, xy = TRUE, na.rm = TRUE)
precip_df <- as.data.frame(precip_annual, xy = TRUE, na.rm = TRUE)

combined_df <- merge(elev_df, precip_df, by = c("x", "y"))
colnames(combined_df) <- c("longitude", "latitude", "elevation", "precip")
```

#sanity check

```{r}
head(combined_df)
summary(combined_df)
plot(combined_df$elevation, combined_df$precip)
```

#Analyze and visualize

```{r}
ggplot(combined_df, aes(x = elevation, y = precip)) +
  geom_point(alpha = 0.3, color = "blue") +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(title = "Precipitation vs Elevation in Sachsen",
       x = "Elevation (m)", y = "Annual Precipitation (mm)") +
  theme_minimal()

model <- lm(precip ~ elevation, data = combined_df)
summary(model)
```

#ANALYSIS OF THE OUTPUT ABOVE - elevation VS precipation;

The scatterplot shows a strong positive relationship between elevation and annual precipitation in Sachsen. As elevation increases, precipitation tends to rise as well. The red regression line indicates a clear upward trend, suggesting that higher-altitude areas receive more rainfall — likely due to orographic effects. While there is some spread in the data at mid elevations (500–700 m), the overall correlation appears strong and linear.


#Statistical Relationship: Precipitation ~ Elevation
```{r}
model <- lm(precip ~ elevation, data = combined_df)
summary(model)
```

#Summary of the statistical relationship result above;

There is a strong and statistically significant positive relationship between elevation and annual precipitation in Sachsen. The model explains approximately 82% of the variation in precipitation. On average, annual precipitation increases by 0.46 mm for every 1 meter increase in elevation. This supports the hypothesis that elevation plays a major role in local climate conditions within the region.


#Make a nice map that compares elevation in Saxony to annual precipation

```{r}
# Convert raster data to data frames
elev_plot_df <- as.data.frame(elev_sachsen, xy = TRUE, na.rm = TRUE)
colnames(elev_plot_df) <- c("longitude", "latitude", "elevation")

precip_plot_df <- as.data.frame(precip_annual, xy = TRUE, na.rm = TRUE)
colnames(precip_plot_df) <- c("longitude", "latitude", "precip")

# Elevation map
elev_map <- ggplot() +
  geom_raster(data = elev_plot_df, aes(x = longitude, y = latitude, fill = elevation)) +
  geom_sf(data = Sachsen, fill = NA, color = "black", size = 0.3) +
  scale_fill_gradient(name = "Elevation (m)", low = "beige", high = "darkgreen") +
  labs(title = "Elevation in Sachsen") +
  coord_sf() +
  theme_minimal()

# Precipitation map
precip_map <- ggplot() +
  geom_raster(data = precip_plot_df, aes(x = longitude, y = latitude, fill = precip)) +
  geom_sf(data = Sachsen, fill = NA, color = "black", size = 0.3) +
  scale_fill_gradient(name = "Precipitation (mm)", low = "lightblue", high = "darkblue") +
  labs(title = "Annual Precipitation in Sachsen") +
  coord_sf() +
  theme_minimal()

# Combine maps side by side
elev_map
precip_map
```